Grappa: Scalable Computing on Mass-Market Hardware for Irregular, Low-Locality Applications

### 2012-07-22 doc/asplos13/abstract.tex
Pass on paper, less related work and results

Grappa is a latency tolerant runtime system for commodity multinode multicore
platforms that addresses the challenges posed in scaling irregular parallel
applications to large datasets. These applications are of growing importance
as interactive analysis of large unstructured data sets emerges as a
commercial and government priority. Grappa serves both as a C++ user library
and as a foundation upon which higher level languages can be developed or
adapted to support development of irregular parallel applications. Grappa
tolerates the delays to distant memory by multiplexing hundreds to thousands
of lightweight tasks to each processor core; balances load via fine-grained
workstealing; and provides efficient synchronization and consistency across
its single shared address space by delegating all accesses to any given global
location to a corresponding processor core. We present a detailed description
of the Grappa system, programming examples using the library interface, and
compare performance for several irregular benchmarks to the Cray XMT, a custom
system used to target the real time graph analytics market.

>>>>

Grappa is a latency tolerant runtime system for commodity multinode multicore
platforms that addresses the challenges posed in scaling irregular parallel
applications to large datasets. These applications are of growing importance
as interactive analysis of large unstructured data sets emerges as a
commercial and government priority. Grappa serves both as a C++ user library
and as a foundation upon which higher level languages can be developed or
adapted to support development of irregular parallel applications. Grappa
tolerates the delays to distant memory by multiplexing hundreds to thousands
of lightweight tasks to each processor core; balances load via fine-grained
workstealing; and provides efficient synchronization and consistency across
its single shared address space. We present a detailed description
of the Grappa system, programming examples using the library interface, and
compare performance for several irregular benchmarks to the Cray XMT, a custom
system used to target the real time graph analytics market.

### 2012-07-22 doc/asplos13/implementation.tex
Pass on paper, less related work and results

Grappa implements a software distributed shared memory. This memory is exposed
through two distinct, non-overlapping address spaces; one optimized for global
access to node-local data, and another optimized for high aggregate random
access bandwidth to low locality data.

All global addresses in Grappa are 64-bit
values. Each byte of global memory is associated with a {\em home core}; this
core is responsible for allocating and modifying that memory. Since today's
commodity processors support a 48-bit virtual address space \TODO{cite?},
Grappa uses the spare bits to distinguish which address space a global address
belongs to, and to represent the home core of a global address.

Any node-local data can be made accessible by other nodes in the system by
wrapping the address and node ID into a 2D global address. This address can
then be accessed with delegate and can also be cached by other nodes. At the
destination the address is converted into a canonical x86 address by replacing
the upper bits with the sign-extended upper bit of the virtual address.

All linear addresses refer to data allocated from a global heap. The
heap metadata is stored on a single node. Currently all heap
operations serialize through this node; while this has been sufficient
for our benchmarks, parallel performance can be achieved through
combining \TODO{ cite MAMA and flat combining}. 2D addresses may refer
to memory allocated from a single processes' heap or from a task's
stack. Figure~\ref{fig:memory-structure} shows how 2D and linear
addresses can refer to other cores' memory.

All global memory operations are implemented with active
messages. Requesters construct a descriptor holding the current task
ID and space for a response, issue their request active message to the
node given by the request address, and suspend themselves. When the
request arrives at the address' home node, the operation is performed
and a response active message is sent. When the response arrives at
the requesting node, the descriptor is filled in with the results of
the operation and the requesting task is woken.

\paragraph{Explicit caches} Grappa cache operations are used to exploit
spatial or temporal locality. They allow a single task to buffer a range of
data, read or modify it, and write the modified version back to its home node.
All fetch and update operations of cache data are expressed explicitly by the
programmer. Cache operations may operate on data that spans multiple blocks of
the linear address space; the programmer provides the starting address and
length, and the cache issues as many active messages as necessary to retrieve
all necessary data.

\subsection{Tasking}

@@

Context switching in Grappa is non-preemptive. Any Grappa routine that
does a potentially-long-latency operation calls a context-switch
routine that gives control of the core back to the scheduler. Users
may also explicit yield control to the scheduler.

As with other cooperative multithreading systems, we treat context
switches as function calls, saving and restoring only the callee-saved
state as specified in the x86-64 ABI \TODO{cite}. This involves saving
six general-purpose 64-bit registers and the stack pointer, as well as
the 16-bit x87 floating point control word and the SSE context/status
register. Thus, the minimum amount of state a cooperative context
switch routine must save according to the ABI is 62 bytes.

Since the compiler sees all calls to the context switch routine, we
can save even less state. Our context switch routine appears to the
compiler as inline assembly; we declare all the registers we need
to save as ``clobbered'' by the inline assembly routine, and the
compiler will issue its own save and restore code as needed. This allows the
compiler to avoid saving any registers that are not used, or are used
for temporary values that are not needed after the context switch.
\comment{I highly doubt this is important in real apps beyond our kernels.  -M
Why not? liveness is a common compiler analysis -BM

@@

Each scheduler has three queues:
\begin{description}
\item{\bf Ready worker queue} This is a FIFO queue of tasks that are
  matched with workers and are ready to execute.
\item{\bf Private task queue} This is a FIFO queue of tasks that must run on this core.
\item{\bf Public task queue} This is a LIFO queue of tasks that are
  waiting to be matched with workers. It is a local partition of a shared
  task pool; this is described in the next section.
\end{description}

Whenever a task yields or suspends, the scheduler makes a decision
about what to do next. First, it determines if
the communication resources should be serviced. This is done on a
periodic basis. Second, it determines if any workers with running
tasks are ready to execute; if so, one is scheduled. Finally, if there
are no workers ready to run, but there are tasks waiting to be matched
with workers, an idle worker is woken (or a new worker is spawned),
matched with a task, and scheduled.

The scheduler uses its core's timestamp counter to track time. The
timestamp counter is read once per scheduling decision. Timestamp
values are not shared between cores. We use timestamp values only to
ensure periodic execution of communication maintenance tasks; being
wrong will only change the rate of these events with a potential
performance cost, but does not cause a correctness violation. Nevertheless, we
assume Grappa will run on machines with modern timestamp counter
implementations that are monotonic and frequency-compensated.

\subsubsection{Work stealing} \label{subsec:implementation-worksteal}
When a \checkme{grappa-process} runs
out of work, it becomes a ``thief'' and asks another other
\checkme{grappa-process} (``victim'') for work. If the victim has tasks
to spare in its public queue, they will be transferred to the thief's public queue.

Most study of work stealing assumes that a processor has one worker
thread and so it only steals when utilization would go to zero. 
Grappa, on the other hand, relies on having many worker threads per core. The
core would be fully-utilized if there is always something on the ready
queue. Even if there are many active tasks, if they are all suspended
for long-latency network requests, then the core is underutilized.
So, there is a choice: should the core use some of this idle time to
perform work stealing or should it just wait as it is possible that local tasks will
create more work. We choose to have a core initiate a steal when all
of the following conditions hold: no workers are ready to run, 
the unstarted task queues are empty, and there are no outstanding steal requests. 
Having only one outstanding steal request throttles steals to
prevent flooding the network, but other scalable quieting mechanisms
would be possible, such as a voting tree\cite{scalableWorkStealingOrCilk98}
or lifelines \cite{lifelines}. Termination detection is not
built into the Grappa task scheduler, rather, it is considered a programming error for
the program to exit without syncing all tasks.

\TODO{where to put the following related}
Fork join computations: implement with futures or continuation passing
style
feed-forward computations: implement with scalable task counting (we
focus on this style in our benchmarks)
(briefly describe global task joiner)

choices we've made
-random victims
-steal from FIFO side of queue
-steal when at the moment readyQ empty and task queues empty
-steal min(chunk\_size, half)

Recent work has explored scalable work stealing in terms of efficient
termination detection.

\subsection{Communication}

@@

At the upper layer, Grappa implements asynchronous active messages
\TODO{cite vonEicken92}. Each message consists of a function pointer, an
optional argument payload, and an optional data payload. When a task
sends a message, the message is copied to a send queue associated with
the message's destination and the task continues execution.

Grappa's lower networking layer aggregates the upper layer's messages
to improve performance. Commodity networks including Infiniband
achieves their peak injection bandwidth \emph{only} when the packet
sizes are relatively large---on the order of multiple kilobytes. The
reason for this discrepancy is the combination of overheads associated
with handling each packet (in terms of bytes that form the actual
Infiniband packet, processing time at the card and processing on the
CPU within the driver stack). Our measurements confirm manufacturers
published data~\cite{infinibandbandwidth}, that with this packet size
the bisection bandwidth is only a small fraction, less than
~\checkme{5\%} of the peak bisection bandwidth.

In our experiments the vast majority of requests were smaller than
\checkme{255}
bytes \TODO{more precise? median?}, far too small to make efficient
use of the network. To make the best use of the network, we must
convert our small messages into large ones.
\TODO{reference to aggregation evaluation later in the paper}

Grappa leverages its lightweight context switching to do this
conversion. When a task sends a message, the message is not
immediately sent. Instead, it is stored in a queue associated with its
destination, to be sent later. If the task requires a response, it
suspends itself and the scheduler finds other work to run while the
task is waiting. When the reply arrives, the suspended task is woken.

There are three situations in which a queue of aggregated messages is
sent. First, each queue has a message size threshold, chosen to give
reasonable network performance. If the size in bytes of a queue is
above the threshold, the contents of the queue are sent
immediately. Second, each queue has a wait time threshold. If the
oldest message in a queue has been waiting longer than this threshold,
the contents of the queue are sent immediately, even if the queue size
is lower than the message size threshold.  Third, queues may be
explicitly flushed in situations where the programmer wants to
minimize the latency of a message at the cost of bandwidth
utilization.

@@

Underneath the aggregation layer, Grappa uses the \gasnet~communication
library \TODO{cite GASNet} to actually move data. All interprocess
communication, whether on or off a cluster node, is handled by the
\gasnet~library. \gasnet~is able to take advantage of many communication
mechanisms, including Ethernet and Infiniband between nodes, as well as
shared memory within a node.

Some networks provide access to a remote machine's memory
directly. This would seem to be a good fit for a programming model
focused on global shared memory, but in fact we do not use it. RDMA
operations are subject to the same message rate limitations as all
other messages on these cards, and thus using raw RDMA operations for
our small messages would make inefficient use of bandwidth. Instead,
we implement remote memory operations with active messages. A byproduct of this design decision is that Grappa is not limited
to RDMA-capable networks.

>>>>

\subsection{Tasks}

@@

Context switching in Grappa is non-preemptive. As with other cooperative
multithreading systems, we treat context switches as function calls,
saving and restoring only the callee-saved state as specified in the
x86-64 ABI \cite{someone}. This involves saving six general-purpose
64-bit registers and the stack pointer, as well as the 16-bit x87
floating point control word and the SSE context/status register. Thus,
the minimum amount of state a cooperative context switch routine must
save according to the ABI is 62 bytes.

Since the compiler sees all calls to the context switch routine, we
can save even less state. Our context switch routine appears to the
compiler as inline assembly; we declare all the registers we need
to save as ``clobbered'' by the inline assembly routine, and the
compiler will issue its own save and restore code as needed. This allows the
compiler to avoid saving any registers that are not used, or are used
for temporary values that are not needed after the context switch.
\comment{I highly doubt this is important in real apps beyond our kernels.  -M
Why not? liveness is a common compiler analysis -BM
Because it has to be live across the entire application.  All real applications do some form of floating point, however minor... -M

@@

Each scheduler has three queues:
\begin{description}
\item{\bf Ready worker queue} This is a FIFO queue of tasks that are
  matched with workers and are ready to execute.
\item{\bf Private task queue} This is a FIFO queue of tasks that must run on this core.
\item{\bf Public task queue} This is a LIFO queue of tasks that are
  waiting to be matched with workers. It is a local partition of a shared
  task pool.
\end{description}

Whenever a task yields or suspends, the scheduler makes a decision about
what to do next. First, it determines if the communication resources
should be serviced. This is done on a periodic basis. Second, it
determines if any workers with running tasks are ready to execute; if
so, one is scheduled. Finally, if there are no workers ready to run, but
there are tasks waiting to be matched with workers, an idle worker is
woken (or a new worker is spawned), matched with a task, and scheduled.
The scheduler ensures that communication maintenance tasks are executed periodically.

\TODO{Is there an off by 1 error in here?  Need to read the AMD64 docs carefully to see if we can sign extend the top bit of the 48 bits of real virtual address space, or if we need to maintain bit 49.  -M}

Grappa implements a software distributed shared memory. All global
addresses in Grappa are 64-bit values. Each byte of global memory is
associated with a {\em home core}; this core is responsible for
allocating and modifying that memory. Since today's commodity processors
support a 48-bit virtual address space \cite{AMD64}, Grappa uses the
spare bits to distinguish which address space a global address belongs
to, and to represent the home core of a global address.

Global memory in Grappa is exposed through two distinct, non-overlapping
address spaces; \textbf{2D global addresses}, optimized for global
access to node-local data, and \textbf{linear global addresses},
optimized for high aggregate random access bandwidth to low locality
data.

Any node-local data can be made accessible by other nodes in the system
by wrapping the address and node ID into a 2D global address. This
address can then be accessed with a delegate and can also be cached by
other nodes. At the destination the address is converted into a
canonical x86 address by replacing the upper bits with the sign-extended
upper bit of the virtual address. 2D addresses may refer to memory
allocated from a single processes' heap or from a task's stack.
Figure~\ref{fig:memory-structure} shows how 2D and linear addresses can
refer to other cores' memory.

All linear addresses refer to data allocated from a global heap. The
heap metadata is stored on a single node. Currently all heap operations
serialize through this node; while this has been sufficient for our
benchmarks, in the future Grappa will provide parallel performance
through combining~\cite{MAMA,flatcombining}.

All global memory operations are implemented with active
messages. Requesters construct a descriptor holding the current task
ID and space for a response, issue their request active message to the
node given by the request address, and suspend themselves. When the
request arrives at the address' home node, the operation is performed
and a response active message is sent. When the response arrives at
the requesting node, the descriptor is filled in with the results of
the operation and the requesting task is placed back on the ready queue.

\paragraph{Explicit caches} Grappa cache operations are used to exploit
spatial or temporal locality. They allow a single task to buffer a range
of data, read or modify it, and write the modified version back to its
home node(s). All fetch and update operations of cache data are
expressed explicitly by the programmer. Cache operations may operate on
data that spans multiple blocks of the linear address space; the
programmer provides the starting address and length, and the cache
issues as many active messages as necessary to retrieve all necessary
data.

\subsection{Load Balancing and Network Aggregation}

\subsubsection{Work stealing} \label{subsec:implementation-worksteal}

When a \checkme{grappa-process} runs out of work, it becomes a ``thief''
and asks another other \checkme{grappa-process} (``victim'') for work.
If the victim has tasks to spare in its public queue, they will be
transferred to the thief's public queue.

Most prior work on work stealing assumes that a processor has one worker
thread and so it only steals when utilization would go to zero. Grappa,
on the other hand, relies on having many worker threads per core. The
core would be fully-utilized if there is always something on the ready
queue. Even if there are many active tasks, if they are all suspended
for long-latency network requests, then the core is underutilized. So,
there is a choice: should the core use some of this idle time to perform
work stealing or should it just wait as it is possible that local tasks
will create more work. We choose to have a core initiate a steal when
all of the following conditions hold: no workers are ready to run, the
unstarted task queues are empty, and there are no outstanding steal
requests. Having only one outstanding steal request throttles steals to
prevent flooding the network, but other scalable quieting mechanisms
would be possible, such as a voting
tree\cite{scalableWorkStealingOrCilk98} or lifelines \cite{lifelines}.
Termination detection is not built into the Grappa task scheduler,
rather, it is considered a programming error for the program to exit
without syncing all tasks.

\TODO{Someone check this paragraph to make sure it is true. :-)  -Mark}

When a grappa node decides to issue a steal request it chooses the
victim node at random.  It also requests a block of tasks at once.  If
this block exceeds the amount of stealable tasks on the victim node,
half of the stealable tasks are stolen instead.  While we did study the
number of tasks to steal at once, we found performance was maximized
with \checkme{128} tasks with little sensitivity around this number,
hence we omit results for this from Section~\ref{sec:results} and simply
use \checkme{128} as the parameter for all experiments.

\comment{
where to put the following related? -BM

Fork join computations: implement with futures or continuation passing
style
feed-forward computations: implement with scalable task counting (we
focus on this style in our benchmarks)
(briefly describe global task joiner)

choices we've made
done -random victims
hmm..  -steal from FIFO side of queue
hmmm... -steal when at the moment readyQ empty and task queues empty
done -steal min(chunk\_size, half)

omit - Recent work has explored scalable work stealing in terms of efficient
termination detection. 
}

\subsubsection{Network Aggregation}

@@

At the upper layer, Grappa implements asynchronous active messages
\cite{vonEicken92}. Each message consists of a function pointer, an
optional argument payload, and an optional data payload. When a task
sends a message, the message is copied to a send queue associated with
the message's destination and the task continues execution.

Grappa's lower networking layer aggregates the upper layer's messages
to improve performance. Commodity networks including infiniband
achieves their peak bisection bandwidth \emph{only} when the packet
sizes are relatively large---on the order of multiple kilobytes. The
reason for this discrepancy is the combination of overheads associated
with handling each packet (in terms of bytes that form the actual
packet, processing time at the card and processing on the
CPU within the driver stack). Our measurements confirm manufacturers
published data~\cite{infinibandbandwidth}, that with this packet size
the bisection bandwidth is only a small fraction, less than
~\checkme{5\%} of the peak bisection bandwidth.

In our experiments the vast majority of requests were smaller than
\checkme{255} bytes \TODO{more precise? median?}, far too small to make
efficient use of the network. To make the best use of the network, we
must convert our small messages into large ones
(Section~\ref{sec:results-aggregation}).  When a task sends a message,
it is not immediately sent, but rather placed in a queue specific to the
destination.

There are three situations in which a queue of aggregated messages is
sent. First, each queue has a message size threshold, chosen to give
reasonable network performance (\checkme{2048 bytes}). If the size in
bytes of a queue is above the threshold, the contents of the queue are
sent immediately. Second, each queue has a wait time threshold
(\checkme{200ms}). If the oldest message in a queue has been waiting
longer than this threshold, the contents of the queue are sent
immediately, even if the queue size is lower than the message size
threshold.  Third, queues may be explicitly flushed in situations where
the programmer wants to minimize the latency of a message at the cost of
bandwidth utilization.

@@

Underneath the aggregation layer, Grappa uses the \gasnet~communication
library~\cite{gasnet} to actually move data. All interprocess
communication, whether on or off a cluster node, is handled by the
\gasnet~library. \gasnet~is able to take advantage of many communication
mechanisms, including ethernet and infiniband between nodes, as well as
shared memory within a node.

Some networks provide access to a remote machine's memory directly. This
would seem to be a good fit for a programming model focused on global
shared memory, but in fact we do not use it. In our experiments, we
found that RDMA operations are subject to the same message rate
limitations as all other messages on these cards, and thus using raw
RDMA operations for our small messages would make inefficient use of
bandwidth. Instead, we implement remote memory operations with active
messages. A byproduct of this design decision is that Grappa is not
limited to RDMA-capable networks.
